---
title: "Fecundity Assay Practice"
output: html_document
date: "2025-07-11"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goals

Visually represent aphid mortality and fecundity

# Load Libraries

```{r load-libraries}
pacman::p_load(readxl, ggplot2, dplyr, tidyr, tidyverse, devtools, scales, install=FALSE)
```

# Load Data

```{r load-data}
aphid_count <- read_excel("data/fecundity_assay_count.xlsx")
## filter for block 2 only 
aphid_block2 <- aphid_count %>%
  filter(block=="2")
```

# Calculations
Calculating how many aphids are in each group as a new data frame
```{r death_count}
aphids_alive_r1_summary <- aphid_block2 %>%
  mutate(
    nymphs_r1 = as.numeric(nymphs_r1),
    alate_alive_r1 = ifelse(alate_alive_r1 == "Yes", 1, 0),
    aphids_alive_r1 = nymphs_r1 + alate_alive_r1
  ) %>%
  group_by(treatment) %>%
  summarise(total_aphids_alive_r1 = sum(aphids_alive_r1, na.rm = TRUE))
```

# Filter for block 2 and treatments of interest
```{r filter}
aphid_block2 <- aphid_count %>%
  filter(block == "2")

aphid_survival <- aphid_block2 %>%
  filter(treatment %in% c("220", "MgCl2")) %>%
  mutate(
    survival_status = case_when(
      alate_alive_r1 == "Yes" ~ "Alive",
      alate_alive_r1 == "No" ~ "Dead",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(survival_status))
```

# Summarize counts and calculate % alive
```{r summarize}
aphid_survival_summary <- aphid_survival %>%
  group_by(treatment, survival_status) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(treatment) %>%
  mutate(percentage = count / sum(count) * 100)
```

# Filter only alive aphids for the plot
```{r filter_alive}
aphid_alive_summary <- aphid_survival_summary %>%
  filter(survival_status == "Alive")
```

# Plot survival
```{r plot}
ggplot(aphid_alive_summary, aes(x = treatment, y = percentage, fill = treatment)) +
  geom_col() +
  labs(
    title = "Aphid Survival - Block 2 - Round 1",
    x = "Treatment", y = "Percentage Alive"
  ) +
  scale_y_continuous(
    limits = c(0, 100),
    labels = percent_format(scale = 1),
    expand = c(0, 0)
  ) +
  scale_fill_manual(values = c("220" = "darkgreen", "MgCl2" = "darkred")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```
# Reshape data: gather r1 and r2 survival into long format
```{r reshape}
aphid_long <- aphid_block2 %>%
  pivot_longer(
    cols = c(alate_alive_r1, alate_alive_r2),
    names_to = "round",
    values_to = "survival"
  ) %>%
  mutate(
    round = ifelse(round == "alate_alive_r1", "Round 1", "Round 2"),
    survival_status = case_when(
      survival == "Yes" ~ "Alive",
      survival == "No" ~ "Dead",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(survival_status))
```

# Summarize percent survival per treatment and round
```{r percent_survival_b2}
aphid_survival_summary <- aphid_long %>%
  group_by(treatment, round, survival_status) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(treatment, round) %>%
  mutate(percentage = count / sum(count) * 100)
aphid_alive_summary <- aphid_survival_summary %>%
  filter(survival_status == "Alive")
```

# Plot survival across both rounds
```{r plot_both_rounds}
ggplot(aphid_alive_summary, aes(x = treatment, y = percentage, fill = treatment)) +
  geom_col(position = "dodge") +
  facet_wrap(~round) +
  labs(
    title = "Percentage of Surviving Aphids by Treatment (Block 2)",
    x = "Treatment", y = "Percentage Alive"
  ) +
  scale_y_continuous(
    limits = c(0, 100),
    labels = percent_format(scale = 1),
    expand = c(0, 0)
  ) +
  scale_fill_manual(values = c("220" = "#005200", "MgCl2" = "#602")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```